---
title: "Effect of Environmental Predictors on horn sizes of male and female Chamois in the Eastern Italian Alps"
author: "Franz Johann, Liangwen He, Matthias Theobald"
date: "18. December 2015"
output: 
  html_document: 
    keep_md: yes
    number_sections: yes
    toc: yes
---

```{r echo=F, message=FALSE}
library(knitr)
library(mgcv)
library(Hmisc)
library(psych)
```

# Introduction



## About the data
We have received a data set of 2679 observations from hunted animals of the species *Rupicapra rupicapra* (Chamois) in the Eastern Italian Alps in the region of Alto Adige (South Tirol). The shot animals are yearlings culled during the hunting season of their second year. The hunting season is from September to December. The data measured from each individual is 

- horn length in mm,
- weight in kg,
- sex,
- Year and Julian Day of the culling, as well as
- Council and Area Code.


Also included in the data set is the council where the animal was shot, which in turn belongs to one of five administrative areas. Furthermore, there are environmental data, aggregated at either council or area level.

## Description of Environmental Data

Because of problems with the data we have one value for temperature, precipitation and snow for each season covered by the life history of the killed individual for the study area. There is also information about the global climate scale North Atlantic Oscillation(NAO) index.

At area level there is substrate with two categories: siliceous and calcareous. Animal density is supplied per year which is then either confirmed in the next year or changed.

We have the geographic coordinate, elevation and aspect for each council. Also recorded is the Percentage of open area in the council. The Normalized Difference Vegetation Index (NDVI) is used as an indicator for food supply.

## Struggling with Data

The first data set we received only included partial weather data. First, we wanted to know about the weather conditions of the winter during the pregnancy of the mother (first winter) but we only had data about the weather of the winter the yearling experienced (second winter).
Second, there was no weather data about the autumn the yearling experienced.
Third, while exploring the data, we found several inconsistencies in the weather data.

Because of that, we asked for a larger data set including the weather data of the first winter during pregnancy and also asked for clarification on the issues with the weather station. We received a second data set where the requested information was provided, but 1100 observations of shot animals were removed, so only 1560 were available. The researchers responsible for the data argued that the removed data was probably not reliable. Our argument though is that each and every observation is useful and should be included in the model. We therefore set out to work on merging the two data sets and filling the holes in the first set with meaningful weather data from the second.

We received information from the curators of the data about the weather stations they used for the different areas. the When trying to bring the two data sets together we found that the relationships between the weather data of different areas are unsteady, although they should have shared the same stations. It turned out that since not all stations were working consistently throughout the study some areas had to borrow weather data from other area. This is a serious problem, we can now no more maintain that the weather data for each council is true.
To deal with this problem, we try to find out one council sample, which on one side has insisted with the weather data from one station continuously from 2005 to 2012, and on the other side can be representative of the whole research area considering its elevation(finally council 39). Then we generalizing its weather data to the whole research area for each year. At last we would introduce an interaction between elevation and aspect, which should reflect the true variance of weather in different councils.

## Loading the data sets into R

```{r loading data}
# helper function: inverse of intersect
outersect <- function(x, y) {
    sort(c(setdiff(x, y),
        setdiff(y, x)))
}

# read data
db_chamois1 <- read.csv("chamois1.csv", sep=";")
db_chamois2 <- read.csv("chamois2.csv", sep=";")


#rename columns so merging can work its magic
# in the new data set, a "2"" is appended at the end of the columns names
db_chamois1$Snow_cover_winter2 <- db_chamois1$Snow_cover_winter
db_chamois1$snow_winter2 <- db_chamois1$snow_winter
db_chamois1$twinter.max2 <- db_chamois1$twinter.max
db_chamois1$twinter.mean2 <- db_chamois1$twinter.mean
db_chamois1$twinter.min2 <- db_chamois1$twinter.min

# delete the old columns
db_chamois1$Snow_cover_winter <- NULL
db_chamois1$snow_winter <- NULL
db_chamois1$twinter.max <- NULL
db_chamois1$twinter.mean <- NULL
db_chamois1$twinter.min <- NULL

# delete columns excel has added
db_chamois2$X <- NULL
db_chamois2$X.1 <- NULL
db_chamois2$X.2 <- NULL
```

After loading the data we face the problem that the second data set has less observations than the first. 

```{r number of observations}
# number of observations in each data set
nrow(db_chamois1); nrow(db_chamois2)
```

There is evironmental data, especially weather data in the second data set that we want to use for all observations of the first.

First we merge the two sets, keeping all observations.
```{r merging data}
# merge the databases
db_merge <- merge(db_chamois1, db_chamois2, all.x=T)

# delete all unnecessary columns
drops <- c(
    "data", # date as string
    "exc.date2", # excel numeric date
    "Julia.date", # year + Julian day
    "date", # year + Julian day + random number
    "x", # coordniates + small random step for spatial analysis
    "y", # same
    "x2", # same
    "y2", # same
    "y_r", #same
    "x_r", #same
    "NS", # North-south facing component of aspect
    "EO", # East-West facing component of aspect
    "Nweight", # normalized weight
    "Nhorn", # normalized horn length
    "h_w", # ratio hor length to weight
    "index" # Indice of above ratio
)

# drop unnecessary columns
db_merge <- db_merge[, !names(db_merge) %in% drops]

# order the dataframe by id
db_merge <- db_merge[order(db_merge$id), ]
```

These are the columns present in the new merged data.

```{r db_merge names, echo=F}
names(db_merge)
```

These 15 columns are not present in the first first set.

```{r outersect table, echo=F}
# what columns are new in the new data set?
outersect(colnames(db_chamois1), colnames(db_chamois2))
```

Our first approach was to take the weather data from the areas present in the second data (area 1 and 6) set and put them in the same area of the first set. We then wanted to interpolate over elevation to get new values for those areas not in the second set (area 2, 3, and 4).

```{r areas in the first and second set}
unique(db_chamois1$area_cod)
unique(db_chamois2$area_cod)
```

We need to take values from the second set and include them in the first set. Here the problem is that the use of weather stations is inconsistent, as well as some of the stations are not present in all councils and some councils use weather data from a station which is unsuitable to represent the whole study area.


```{r fixing weather}

# make weather data consistent
# select all weather data
weather_data <- c(
    "snow_winter1",
    "Snow_cover_winter1",
    "r_apr_mag_1",
    "r_giu_lug_1",
    "r_ago_set_1",
    "r_spring1",
    "r_newsummer1",
    "r_autumn",
    "snow_winter2",
    "Snow_cover_winter2",
    "r_apr_mag_2",
    "r_giu_lug_2",
    "r_ago_set_2",
    "r_spring2",
    "r_newsummer2",
    "twinter.min1",
    "twinter.max1",
    "twinter.mean1",
    "tspring1.min",
    "tspring1.max",
    "tspring1.mean",
    "tsummer1.min",
    "tsummer1.max",
    "tsummer1.mean",
    "tautumn.min",
    "tautumn.max",
    "tautumn.mean",
    "twinter.min2",
    "twinter.max2",
    "twinter.mean2",
    "tspring2.min",
    "tspring2.max",
    "tspring2.mean",
    "tsummer2.min",
    "tsummer2.max",
    "tsummer2.mean"
)
```

Now, we first have to check in which councils chamois were hunted each year.
```{r consistency check}
# consistency of kills in councils: not every year a chamois was shot in every council
with(db_merge, tapply(horn, list(council_cod, year), length))
# we need the consistent council to select which weather data to use
```

Councils 27, 10, 8 39 and 49 are ok and use weather data from statsions with continuous measurements throughout the study.

The data from council 39 and 49 are from the weather station *Pinzolo Val Genova*, which we deem best for representing the wohle study area. We picked council 39 for our purpose.

```{r weather stations}
# weather data from council 27 for each year
station27 <- unique(db_merge[db_merge$council_cod==27, c("year", weather_data)]) # not okay: year2007
# for the weather data in autumn, council27 has 2 values for the year 2007
station10 <- unique(db_merge[db_merge$council_cod==10, c("year", weather_data)])
station8 <- unique(db_merge[db_merge$council_cod==8, c("year", weather_data)])
# the weather stations used in council 8 and 10 are not representative for the whole area

station39 <- unique(db_merge[db_merge$council_cod==39, c("year", weather_data)]) # ok, snow, T, and Prec are from the higher stations both
station49 <- unique(db_merge[db_merge$council_cod==49, c("year", weather_data)]) # same as 39
# representative!


# merge the new weather data set into the old one
# removing the old weather columns first
db <- merge(db_merge[, !names(db_merge) %in% weather_data], station39, by="year", all=F)
```

Lastly, we remove columns of variables which are either updated or removed in the new data set because they have errors.

```{r remove columns}
# remove the old summer precipitation data
db$r_summer_1 <- NULL
db$r_summer_2 <- NULL

# remove old may ndvi. it's incorrect
db$ndvi.may1 <- NULL
db$ndvi.may2 <- NULL
```

A quick check for NA in the new data set:
```{r check for NA}
# check for NA
with(db, tapply(snow_winter1, year, function(y) sum(is.na(y))))
summary(db$snow_winter2)
summary(db$r_autumn)
summary(db$tautumn.min)
# No NAs!
```


We now have a data set with `r nrow(db)` observations and `r ncol(db) - 1` variables as potential predictors.


# Methods and Results

## Data Exploration


```{r sex as factor, echo=F}
# sex as factor
db$f.sex <- as.factor(db$sex)
levels(db$f.sex) <- c("female", "male") # relevel sex
summary(db$f.sex)
```

Of the `r nrow(db)` animals, 1218 are female and 1461 are male.

### Horn length

As our dependent variable, we wanted to know how horn length varies between sexes.

First, we wanted to see the distribution of Horn size.
```{r horn distr}
summary(db$horn)
hist(db$horn, breaks=40, freq=FALSE, main="Hornlength", xlab= "Length [mm]")
lines(density(db$horn), lwd=1.5,lty =2)
```

Our dependent variable semms to be very close to normally distributed. However, we wanted to see whether the same holds true for the horn length between sexes.

Mean horn length of male and female and dfifference between sexes:
```{r horn sex mean}
(horn_sex <- with(db, tapply(horn, f.sex, mean)))
diff(horn_sex)
```

```{r horn length sd}
boxplot(horn~f.sex, data = db, main= "Hornlength ",ylab= "length [mm]", xlab= "sex")
```

There are some data points far outside, but we accept them as still being within natural variability.

```{r}
hist(db$horn[db$sex==1],breaks=40,freq=FALSE,col=rgb(1,0,0,0.1), main="Hornlength", xlab= "Length [mm]")
hist(db$horn[db$sex==2],breaks=40,freq=FALSE,add=TRUE,col=rgb(0,0,1,0.1))
lines(density(db$horn[db$sex==1]), col="red", lwd=2,lty =2)
lines(density(db$horn[db$sex==2]), col="blue", lwd=2, lty =1)
legend("topright",lwd=c(2,2), lty= c(2,1),col=c("red","blue"),legend=c("female", "male"))
```

Also horn lengths between sexes are close to normally distributed.

```{r}
with(db, tapply(horn, f.sex, sd))
```

Males show a higher variability of horn length in general, although the difference is fairly small.

Here we wanted to see how horn lengths differ over the years between the sexes.
```{r horn length year sd}
# change over the years
(horn_sexyear <- with(db, tapply(horn, list(year, f.sex), mean)))
apply(horn_sexyear, 2, summary)
apply(horn_sexyear, 2, sd)
# males show less variance in mean horn length over the years
```

There is some variability over the years that suggests using year as random effect in our model later on. However, males show less variability in horn lengths over the years than females.

```{r}
boxplot(horn~year, dat=db, main = "Hornlength", xlab="Year", ylab= "Hornlength")
# in different years the distribution of hornlength is different
```


### Weight

Weight is measured in kilograms of the empty (i.e. gutted) yearling.

```{r summary weight}
summary(db$weight)
```

We deem an empty weight of more than 25 kg as too heavy for a yearling. Their mean weight is at around 15 kg. We therefore removed five individuals heavier than 25 kg from the data set.

```{r remove too heavy animals}
nrow(subset(db, weight > 25))
db <- subset(db, weight < 25)
```

Distribution of weight:
```{r weight distr}
summary(db$weight)
hist(db$weight, breaks=40, freq=F, xlab= "Weight [kg]", main="")
hist(db$weight[db$sex==1],breaks=20,freq=FALSE,col=rgb(1,0,0,0.1), add=T)
hist(db$weight[db$sex==2],breaks=20,freq=FALSE,add=TRUE,col=rgb(0,0,1,0.1))

lines(density(db$weight), lwd=2, lty=3)
lines(density(db$weight[db$sex==1]), col="red", lwd=2, lty = 2)
lines(density(db$weight[db$sex==2]), col="blue", lwd=2, lty=1)
legend("topright",lwd=c(2,2,2),lty= c(1,2,3), col=c("red","blue", "black"),legend=c("female", "male", "both"))
```

Weight has a strange distribution. Plotting the data reveals why.

```{r}
plot(horn ~ weight, data=db, main="Horn ~ Weight", xlab="Weight [kg]", ylab="Horn length [mm]")
```

The plot of horn length against weight shows some anomaly because of different scales of measurement. We analysed the scale detail of the data.

```{r classifiying weight}
# Problems with integers and floats in weight data
# selecting all weight data which are whole numbers
tol <- 1e-12
int_weight <- rep(0, nrow(db))

int_weight[sapply(db$weight, function(y) min(abs(c(y%%1, y%%1-1))) < tol)] <- 1

sum(int_weight) # whole numbers in weight
length(int_weight) - sum(int_weight) # floating point numbers

# are the integers and floating point weigths evenly distributed over the areas?

plot(with(db, tapply(int_weight, area_cod, function(y) sum(y)/length(y) * 100)), type="h", cex=1.4, main="Weight: perentage of integers", xlab="Area", ylab="")
```

In Some areas (or councils) the weight seems to be either guessed or the carcass is weighed on a scale displaying only whole nuembers. Although we thought about reclassifying weight data, we did not have enough time to include that in our models.


Like horn size we wanted to see the difference of weight between sexes.

```{r differences in weight between sexes}
## differences in weight between sexes
(weight_sex <- with(db, tapply(weight, f.sex, mean)))
diff(weight_sex) # difference = 0.3 kg

#graphical
boxplot(weight~f.sex, data = db, ylab= "Weight [kg]", xlab= "sex")
# but males have a higher variance in weight for the individuals
```

The difference in mean weight is not as substantial as the difference in mean horn length between sexes. However, males have a higher  median weight.

```{r change of weight over years}
(weight_sexyear <- with(db, tapply(weight, list(year, f.sex), mean)))
apply(weight_sexyear, 2, summary)
apply(weight_sexyear, 2, sd)

```

As in horn length, males show a smaller variance in mean weight over the years than females.

### Julian Day

Julian Day is the day of year the animal was shot. The hunting season begins in September and ends at the end of the year. It is consistent over the years.

```{r}
range(db$Jday)#[1] 247 364
max(db$Jday) - min(db$Jday)#timespan 117 days
with(db, tapply(Jday, year, min))
with(db, tapply(Jday, year, max))
```

There is an 

```{r}
fgam1 <- gam(weight ~ s(Jday, bs="ts") + s(council_cod, bs="re") + s(year, bs="re"), data=db)
summary(fgam1)
plot(fgam1, select=1)
```
Apparantly the weight increases over the hunting period until ~ day 300 when it rapidly drops. This could  be explained that before October the animals find enough grass for feeding and grwoing, while later the winter has set on and they start to burn their body fat reserves. However, the effect is really small.


```{r}
Julgam1<-gam(weight~s(Jday, bs="ts") + f.sex + s(council_cod, bs="re") + s(year, bs="re"), data=db)
vis.gam(Julgam1, plot.type="persp", theta=-145, main="Weight ~ Julianday + Sex",zlab="weight", xlab="female       male",ylab="Day of Hunting Season")
summary(Julgam1)
```








## effect of Julian day on horn length  male & female #########################

#Loess_horn_Julianday_male------------------------------------------------------------
fmLoess_Jday_m <- loess(male$horn~male$Jday,family="gaussian",span=0.8)
newday_m <- seq(from=min(male$Jday), to=max(male$Jday),by=1)
pred_Horn_m <- predict(fmLoess_Jday_m, newday_m, se=TRUE)
plot(horn~Jday, main ="Loess: Horn~Julianday",ylab="Hornlength [mm]",xlab="Day of the year",pch=1)
lines(pred_Horn_m$fit~newday_m,col="blue", lwd=2)


#Loess horn_Julianday_Female -------------------------------------------------------
fmLoess_Jday_f <- loess(female$horn~female$Jday,family="gaussian",span=0.8)
newday_f <- seq(from=min(female$Jday), to=max(female$Jday),by=1)
pred_Horn_f <- predict(fmLoess_Jday_f, newday_f, se=TRUE)
lines(pred_Horn_f$fit~newday_f,col="red", lwd=2, lty=2)

abline(h=140,col="darkgreen",lwd=2,lty=3)
abline(h=160,col="darkgreen",lwd=2,lty=3)
legend("topright",lwd=c(2,2,2),lty=c(1,2,3),col=c("blue","red","darkgreen"),legend=c("male","female", "fix value 140 mm & 160 mm"),cex=0.8)
# Loess regression shows decreasing hornlength at the beginning of the hunting season and increasing hornlength after day 275

Julgam4<-gam(horn~s(weight)+s(Jday),data=db)
vis.gam(Julgam4,theta=60,main="Hornlength ~ weight+Jday",zlab="Hornlength")

# Gam with Interaction
Julgam5<-gam(horn~s(weight,Jday),data=db)
vis.gam(Julgam5,theta=60,main="Hornlength ~ (weight, Jday)",zlab="Hornlength")
# the decrease could be effect of hunting selection but the effect has to be further invesigated including random effects of council.



### Elevation




### Councils and spatial distributions

We wanted to see the spatial distribution of the councils, as well as the spatial distribution for our main predictors.

```{r spatial distribution of weight}
fm_spatial_weight <- gam(weight ~ s(x.council, y.council), data=db)
vis.gam(fm_spatial_weight, plot.type="contour", main="Weight", xlab="x coords", ylab="y coords")
points(db$x.council, db$y.council, col=db$area_cod, pch=20, cex=1.6)
legend("bottomright", legend=paste("Area", unique(db$area_cod)), col=unique(db$area_cod), pch=20)
```

```{r elevation, unweighted and interpolated!}
# interpolated elevation 
fm_spatial_ele <- gam(q_media~s(x.council, y.council), data=db)
vis.gam(fm_spatial_ele, plot.type = "contour", main="Elevation", xlab="x coords", ylab="y coords")
points(db$x.council, db$y.council, col=db$area_cod, pch=20, cex=1.6)
legend("bottomright", legend=paste("Area", unique(db$area_cod)), col=unique(db$area_cod), pch=20)
```

The spatial distribution suggests that our models might have spatial autocorrelation.

Also, the distribution of councils and aggregation of data at council levels suggests that we should include councils as random factor besides years.

```{r kills in councils}
t(with(db, tapply(horn, council_cod, length)))

```

Council 15 has the most hunted anmimals

```{r}

hist(db$council_cod, breaks=length(unique(db$council_cod)), main="kills per council", xlab="council")

with(db, tapply(horn, list(council_cod, year), length))

```

Also consistently over the years, council 15 has the most kills per year.





