---
title: "randomForest"
output: html_document
---

## RandomForset analyses 
```{r,warning=FALSE,message=FALSE}
library(randomForest)
```
Randomforest analysis of selected variables
```{r,message=FALSE, warnings=FALSE}
both_sexes<-randomForest(horn ~ (sex) + (x.council) + (y.council) + (Jday) + (q_media) + (q_min) + (weight) + f.substrate + (density) + (ndvi.slop1) + (ndvi.maxincr1) + (ndvi.may1.new) + (ndvi.summer1) + (ndvi.summer2) + (ndvi.slop2) + (ndvi.maxincr2) + (Perc.area.aperta) + (snow_winter1) + (aspect) + (r_newsummer1) + (snow_winter2) + (r_spring2) + (r_newsummer2) + (twinter.mean1) + (tspring1.mean) + (tsummer1.mean) + (twinter.mean2) + (tspring2.mean) + (tsummer2.mean), ntree=500, data=db)
varImpPlot(both_sexes)
```

Sex, weight and the day of hunting (Julianday) are by far the most relevant variables.

For a better understanding of the importance of the less relevant factors the variables weight and Julianday are ommitted in the following randomforest analysis and analysis are performed sex separated.

Randomforest for male chamois and plot of variable importance:

```{r}


male<-randomForest(horn ~(x.council) + (y.council) + (q_media) + (q_min) + f.substrate + (density) + (ndvi.slop1) + (ndvi.maxincr1) + (ndvi.may1.new) + (ndvi.summer1) + (ndvi.summer2) + (ndvi.slop2) + (ndvi.maxincr2) + (Perc.area.aperta) + (snow_winter1) + (aspect) + (r_newsummer1) + (snow_winter2) + (r_spring2) + (r_newsummer2) + (twinter.mean1) + (tspring1.mean) + (tsummer1.mean) + (twinter.mean2) + (tspring2.mean) + (tsummer2.mean), ntree=500, data=male_db)
varImpPlot(male)
```

RandomForest for female chamois and plot of variable importance:

```{r}
female<-randomForest(horn ~(x.council) + (y.council) + (q_media) + (q_min) + f.substrate + (density) + (ndvi.slop1) + (ndvi.maxincr1) + (ndvi.may1.new) + (ndvi.summer1) + (ndvi.summer2) + (ndvi.slop2) + (ndvi.maxincr2) + (Perc.area.aperta) + (snow_winter1) + (aspect) + (r_newsummer1) + (snow_winter2) + (r_spring2) + (r_newsummer2) + (twinter.mean1) + (tspring1.mean) + (tsummer1.mean) + (twinter.mean2) + (tspring2.mean) + (tsummer2.mean), ntree=500, data=female_db)
varImpPlot(female)
```


After sex, weight and Julianday the NDVI of summer and May, elevation (q_edia) and pecentage of open area have a high rank in the importance of variables. The rank of the variables differs in the male and female ranfomForest importance analyses.
